A Logic-Based Framework for Digital Signal Processing

### 2008-06-04 trunk/paper/paper.tex

\begin{abstract}
In this paper we describe a logic-based framework for Digital Signal Processing. This framework is implemented on SWI-Prolog and is formed of interfaces with different DSP libraries and APIs extending this logic language ``to speak'' DSP. This approach presents the advantage, in respect of other integrated APIs, of being application implementation independent, so the extended logic language can be loaded an embedded in any application or language interfacing SWI-Prolog. The new DSP capabilities are public as predicates grouped in modules which offer different levels of abstraction. Low-level modules offer predicates to communicate directly with the API. Task-oriented modules define some default routines which specific goals and procedures. Finally we can relate input/output through ``black boxes'' representing a semantic relationship between them. Modules can communicate by using commong semantic data types declared as functors which determine the possible workflows. Such a framework allow us to build complex DSP queries as procedures as well as simplify the extraction of audio processing knowledge with common semantics. Finally, Concurrent Transaction Logic allows us to wrap the procedural meaning of predicate rules into a declarative definition of transactions controlling the execution and updating of the processes.
\keywords{DSP predicates, computation query, computation reasoner, DSP scenarios, RDF}
\end{abstract}

@@

Digital signal processing is a broad field and an object of an important research. Withing this research community, particular investigators are discovering and developing new algorithms for signal processing and analysis. Many projects aim at developing formal APIs providing a framework for algorithms implementation and publication. However, these APIs are normally designed as a closed environment which hardly interacts with other related APIs and applications. Although several APIs are wrapped and exported as ``plugin'' libraries which can be embedded in different applications, they still require the implementation of specific and non-interactive hosts.

@@

In this paper we address this problematic situation by proposing and developing SWI-DSP, a logical framework for \textit{Audio Processing} (as subset of DSP) that extends the SWI-Prolog vocabulary. We rely on a logic programming language with aggregated audio analysis capabilities getting rid of the application-dependency paradigm and offering a single logic front-end which is generally more understandable than a closed implementation in any Procedural or Object Oriented language. In addition to this, we are keen on using logic resources to finally achieve a more capable DSP engine.

In \S 2 we identify the most common audio analysis tasks and summarize several relevant \textit{open source} APIs aimed at performing such tasks. We evaluate the advantages and limitations of these APIs in order to find the requirements for SWI-DSP. In \S 3 we start viewing the system from the top and how it defines an interesting workflow that characterizes input/output relationships between task-oriented modules. This workflow is determined by semantic types, a binary data standard and procedural declarations of DSP rules. \S 4 goes into the implementation of each task-oriented module which are indeed communication layers with dedicated ``plugin'' libraries that are in charge of the actual computation and offer an open development framework.

Thus, SWI-DSP extends SWI-Prolog vocabulary to perform audio analysis and processing by defining semantic types and predicates, triggering external computation performed by dedicated APIs, which are ordered in modules with different levels of abstraction. However, this front-end does not present an efficient and manageable framework as it requires specific SWI-Prolog knowledge to create programs suitable for the concrete needs and it does not offer any efficient tabling mechanism. We tackle this problem in \S 5 by defining DSP ``scenarios'' which offer a semantic view of the system to the user without losing computational flexibility.

We also describe as a Standard, how Concurrent Transaction Logic is the solution to provide an efficient engine allowing a formal data updating mechanism and computation reasoning capabilities. Finally, \S 6 presents related works that use SWI-DSP to create an environment for multimedia data management and visualisation. We conclude in \S 7 by assessing the value and scope of this work and introducing future advances.

\section{Audio Analysis tasks and solutions}\label{sec:tasks}

Audio analysis and synthesis APIs offer formal implementations for development and use of different algorithms and techniques. These algorithms are focused on different tasks that can be classified in these 3 groups:

\begin{itemize}
 \item Processing and Effects
 \item Feature Extraction
 \item Classification
\end{itemize}

\subsection{Audio Processing}\label{subsec:effects}

This tasks group can be characterized by having audio signals as input and output, so an input signal is processed in such a way that we retrieve an altered, processed version of the original signal. Some examples of audio processing tasks are filtering, effects processing and signal generation.

One important dedicated API is the Linux Audio Developer's Simple Plugin API (LADSPA) which has been succesfully used in many systems and applications and is a widely spread implementation for algorithms development. We host LADSPA in SWI-DSP but there are other interesting libraries like DSSI plugins for effects and LV2 (new generation of LADSPA) which are targets for future extensions. (this is too simple i think)

@@

Feature extraction is an important paradigm in signal research as it focuses on extract qualitative characteristics of the signal in a different domain than the input. Features are classified in different levels of abstraction from frequency and time domain characteristics to high-level audio ones.

The Vamp Plugins API offers a standard and utilities to develop feature extraction plugins that can be hosted in any application able to control the ``plugin lifecycle''. We host Vamp plugins as basis for feature extraction in SWI-DSP. In general plugin libraries are characterized for being easy for development but application-dependent. By embedding plugins in SWI-DSP we offer a flexible development framework and make an interactive interface to deal with plugins in several ways.

@@

With the same motivation of this work there are some integrative APIs which combines different tasks to perform low high-level operations in a closed implementation. CLAM, jMIR and Marsyas present a complex framework collecting tools to perform different DSP tasks. The most interesting feature of these frameworks is the possibility to create DSP networks combining specific algorithms and modules thanks to a workflow design.

CLAM also offers a metamodel of DSP processes wich provides semantics to design the networks properly presenting a higher and more manageable level of abstraction. However, these APIs still present two important limitations that we pretend to solve:

\begin{itemize}
 \item There do not offer a common format to exchange data or an easy way to interact with other applications.
 \item They offer a procedural system unable to reach higher-level knowledge by itself.
\end{itemize}

@@

Classification is one the most interesting and complex paradigms in any specific domain exploiting the ``intelligence'' of computational machines. The WEKA API collects different machine-learning algorithms and provides a formal interface (in Java) for development and interaction with other systems.

However, this generic API does not provide semantics to create a dataset which is the most intriguing and complex issue for classifying data properly. Marsyas and jMÍR integrate WEKA orienting it for music classification purporses by offering a code interface with other modules (Marsyas) or by extending the WEKA file format ARFF (jMIR). Both solutions still seem to lack of effective semantics to perform classification of audio data.

\section{SWI-DSP workflow}\label{sec:dspworkflow}

@@

tempo(Input, Tool, Tempo):-
        Input = 'Signal'(Ch, Sr, L, PCMs),
        Tool = 'Vamp',
        vamp_feature_of('tempo', Signal, WholeFeature).
vamp_feature_of(Type, Signal, WholeFeature):-
        select_plugin(Type, PluginKey, Output),
        vamp_outputs_for(Signal, PluginKey, Output, Features),
        flatten(Features, WholeFeature).
vamp_outputs_for(Signak, PluginKey, Output, Feature):-
        vmpl_load_plugin_for(PluginKey, Signal, Plugin),
        set_blockSize(Plugin, BlockSize),
        set_stepSize(Plugin, StepSize),
        get_channels(Signal, Channels),
        vmpl_initialize_plugin(Plugin, Channels, StepSize, BlockSize),
        vamp_compute_feature(Signal, StepSize, BlockSize, Outputs, Plugin, WholeFeature).
vamp_compute_feature(Signal, StepSize, BlockSize, Outputs, Plugin, WholeFeature):-
        get_samples_per_channel(Signal, L),
        findall(Features, vamp_process_signal(Signal, L, StepSize, BlockSize, Output, Plugin, Features), FeatureSet),
        get_sample_rate(Signal, SampleRate),
        vmpl_remaining_features(Plugin, L, SampleRate, Output, Remaining),
        append(FeatureSet, Remaining, RawFeatures),
        delete(RawFeatures, [], WholeFeature).
vamp_process_signal(Signal, L, StepSize, BlockSize, Output, Plugin, Features):-
        set_limit_framing(L, StepSize, Limit),
        set_framing(StepSize, L, Limit, Start),
        vmpl_process_block_framing(Plugin, Signal, Start, BlockSize, Output, FeatureSet).

@@

This module defines any computation which outputs a Signal term. This

The module swiladspa offers communication with LADSPA which is the most generic and spread plugin API for general audio processing.

@@

@@

So far we have presented how SWI-DSP provides a vocabulary of logic predicates to perform audio analysis computations, but this logic front-end is still inappropiate for such tasks. We have organized the modules in communication layers and task-oriented modules, but there is an important trade-off between understandability and specification. The communication layer even offering an application-independent access to the computation libraries still requires a high knowledge of such libraries besides a SWI-Prolog familiarization which is not expected in most part of the potential users. The task-oriented module which is itself very layered provides high-level predicates which are missing some important options at setting-up.

In addition to this, we need require a mechanism to update results and keep track of computation sequences, so that computations are not restricted to one SWI-Prolog session, allowing us to reuse and share them. This section tackles this problem by extending SWI-DSP as an expert system with defined ``DSP scenarios'' and by introducing Concurrent Transaction Logic on top of our extended predicate-logic language.

@@

'Classification'(Identifier, ListOfCategories, ListOfDataRecords, Classifier, Classification) (have to look at it better)

@@

We have presented a powerful reasoner for multimedia analysis implemented as a logic-based framework that aims at converting low-level computations into shareable and distributed knowledge.

>>>>

\begin{abstract}
In this paper we describe a logic-based framework for Digital Signal Processing. This framework is implemented as a set of SWI-Prolog modules developing interfaces to heterogeneous signal processing libraries. A set of functors capturing common datatypes and a standard format for binary data are used to mediate our different analysis components. The system provides the user with a unified and flexible front-end to define signal processing workflows. The framework develops a signal processing-enabled reasoner capable of performing the primary tasks through the interfaces: decoding (specific libraries for each format), feature extraction (Vamp plugins), audio processing and effects (LV2/LADSPA plugins) and classification (WEKA machine learning API).

\keywords{Digital Signal Processing, SWI-prolog modules, signal processing workflows, processing-enabled reasoner, interfaced open source engines}
\end{abstract}

@@

Digital signal processing (DSP) is a broad field and an important object of engineering research. Within this research community, particular investigators are discovering and developing new algorithms for signal processing and analysis. Many projects aim at developing formal APIs providing a framework for algorithms implementation. However, these APIs are normally designed as a closed environment which hardly interacts with other related APIs and applications. Although several APIs are wrapped and exported as ``plugin'' libraries which can be embedded in different applications, they still require the implementation of specific and non-interactive hosts.

@@

In this paper we address this problematic situation by proposing and developing SWI-DSP, a logic-based framework for \textit{Audio Processing} (as subset of DSP) that extends the SWI-Prolog vocabulary. We rely on a logic programming language with aggregated audio analysis capabilities getting rid of the application-dependency paradigm and offering a single logic front-end. Logic resources are also used to achieve a more capable DSP engine. The main and primary motivation for a logic-based implementation of such a system is to turn our conception of traditional procedural applications for computation of data towards semantic frameworks to manage knowledge involving specific computations.

In \S 2 we identify the most common audio analysis tasks and summarize several relevant \textit{open source} APIs aimed at performing such tasks. We evaluate the advantages and limitations of these APIs in order to find the requirements for SWI-DSP. In \S 3 we start viewing the system from the top and how it defines an interesting workflows-based architecture that characterizes input/output relationships between task-oriented modules. These workflows are possible by means of common functors and a binary data standard format. \S 4 goes into the implementation of each task-oriented module that develops communication layers with dedicated audio processing libraries that are in charge of the actual computation and offer an open development framework.

Thus, SWI-DSP extends SWI-Prolog vocabulary to perform audio analysis and processing by defining functors (capturing common datatypes) and predicates, triggering external computation performed by dedicated APIs, which are ordered in modules with different levels of abstraction. However, this architecture does not present an efficient and manageable framework as it requires specific module knowledge to create programs suitable for the concrete needs. We tackle this problem in \S 5 by defining an unified and flexible front-end, composed by DSP ``scenarios'', which allows the user to define workflows with understandable semantics.

Data management is an important paradigm involved in many domains and also in the multimedia research domain. Computations can not often be retrieved from session to session or are stored in local filesystems unlikely directed to extend any shareable source of knowledge because they are not expressed in a standard format with common semantics. This problem also encourages the approach presented in this work, based on a predicate calculus implementation (SWI-Prolog), in two different ways:

\begin{itemize}
 \item Concurrent Transcation Logic can be used to extend the predicate calculus to provide an unified logic framework for computations and data updating (given a database implementation).
 \item SWI-Prolog provides an important set of interfaces to other languages (to build user applications embedding the system) and packages that allow us to interact with semantic web technologies which seem to be the best framework to provide a common source of knowledge.
\end{itemize}

Although SWI-DSP does not develop a data management system itself, we discuss this problem in \S 6 and present related works which contribute to achieve such a multimedia knowledge reasoner/manager. We conclude in \S 7 by assessing the value and scope of this work and introducing future advances.

\section{Audio processing tasks and solutions}\label{sec:tasks}

A particular application of DSP is the study of one-dimensional, time domain signals that in real world often exist as audio signals. Audio signals evolve in time and have acoustic properties produced by the signal structure and shape. Music and speech are the main domains involving audio signals.

The properties of an analog and continuous signal (acoustic signals) can be studied by analysing their digital (sampled and quantified) versions. A one-dimension digital signal is therefore a processable sequence of values in a discretized time domain that represents the audio signal. Digital signals allow machine processing useful to modify the orignal signal or extract characterisitics from it. We can classify the digital signal processing paradigm in these main tasks:

\begin{itemize}
 \item Effects
 \item Feature Extraction
 \item Classification
\end{itemize}

\subsection{Effects}\label{subsec:effects}

A signal may be presented in a raw or non desirable way for its analysis so a basic stage of many systems is the pre-processing. The signal is processed to obtain a satisfying version of it by different techniques (filtering, normalization, amplification...).

It is very common in applied systems to provide tools that modify acoustic properties of the signal by introducing specific effects. Reverberation, delay or distortion are examples of effects that can be digitally obtained reproducing natural or instrumental alterations of the signal.

In practice, we can characterize this group of tasks for its input/output relationships. The operation outputs a signal (which may be time, frequency or multidimensional domain) when is fed with an input signal (accepting several domains as well). Oscillators and synthesis systems outputs signals without input.

One important dedicated API is the Linux Audio Developer's Simple Plugin API (LADSPA) \cite{ladspa}  which has been succesfully used in many systems and applications and is a widely spread implementation for algorithms development. We host LADSPA in SWI-DSP but there are other interesting libraries like DSSI plugins for effects and LV2 (new generation of LADSPA) which are targets for future extensions.

@@

Acoustic properties which are the substract of the psychological perspecitve of audio signals are conveyed in the signal and therefore in their digital representations. The so-called feature extraction paradigm aims at extracting different levels of audio features to characterize, classify and compare the audio signals (and any signal in general). The domain of the feature represents a qualitative side of the signal (tempo, chromatism, timbre...), so features do not convey all the signal information. This is interesting as distinct signals may have some very close features and differ much in others.

The Vamp Plugins API \cite{vamp} offers a framework to develop feature extraction plugins that can be hosted in any application able to control the ``plugin lifecycle''. We host Vamp plugins as basis for feature extraction in SWI-DSP. In general plugin libraries are characterized for being suitable for development but application-dependent.

@@

With the same motivation of this work there are some integrative APIs which combines different tasks to perform low and hig-level operations in a closed implementation. CLAM \cite{clam}, jMIR \cite{jmir} and Marsyas \cite{marsyas} present a complex framework collecting tools to perform different DSP tasks. The most interesting feature of these frameworks is the possibility to create DSP networks combining specific algorithms and modules thanks to a workflow-based architecture (Marsyas and CLAM) or with a descriptive task format (jMIR).

CLAM also offers a metamodel of DSP processes wich provides semantics to design the networks properly presenting a higher and more manageable level of abstraction. However, these APIs still present two important limitations that our work is aimed at solving:

\begin{itemize}
 \item They do not offer a common format to exchange data or an easy way to interact with other applications.
 \item They are based on a procedural system unable to reach higher-level knowledge by itself.
\end{itemize}

@@

Classification is one the most interesting and complex paradigms in any specific domain exploiting the ``intelligence'' of computational machines. WEKA (Waikato Environment for Knowledge Analysis) \cite{weka} collects different machine-learning algorithms and provides several user interfaces to manage the analysis. It aslo provdes a formal interface (in Java) for development and interaction with other systems.

However, this generic API does not provide semantics to create a dataset which is the most intriguing and complex issue for classifying data properly. Marsyas and jMÍR integrate WEKA orienting it for music classification purporses by offering a code interface with other modules (Marsyas) or by extending the WEKA file format ARFF (jMIR). Both solutions still seem to lack of effective semantics and a interactive front-end to perform classification of audio data.

\section{SWI-DSP architecture}\label{sec:architecture}

@@

tempo(Input, Tool, Tempo):-
     Input = 'Signal'(Ch, Sr, L, PCMs),
     Tool = 'Vamp',
     vamp_feature_of('tempo', Signal, WholeFeature).
vamp_feature_of(Type, Signal, WholeFeature):-
     select_plugin(Type, PluginKey, Output),
     vamp_outputs_for(Signal, PluginKey, Output, Features),
     flatten(Features, WholeFeature).
vamp_outputs_for(Signak, PluginKey, Output, Feature):-
     vmpl_load_plugin_for(PluginKey, Signal, Plugin),
     set_blockSize(Plugin, Block),
     set_stepSize(Plugin, Step),
     get_channels(Signal, Channels),
     vmpl_initialize_plugin(Plugin, Channels, StepSize, BlockSize),
     vamp_compute_feature(Signal, Step, Block, Outputs, Plugin, WholeFeature).
vamp_compute_feature(Signal, Step, Block, Outputs, Plugin, Features):-
     get_samples_per_channel(Signal, L),
     findall(F, vamp_process_signal(Signal, L, Step, Block, Output, Plugin, F), FSet),
     get_sample_rate(Signal, SampleRate),
     vmpl_remaining_features(Plugin, L, SampleRate, Outputs, Remaining),
     append(FSet, Remaining, RawFeatures),
     delete(RawFeatures, [], Features).
vamp_process_signal(Signal, L, StepSize, Block, Out, Plugin, Features):-
     set_limit_framing(L, StepSize, Limit),
     set_framing(StepSize, L, Limit, Start),
     vmpl_process_block_framing(Plugin, Signal, Start, Block, Out, Features).

\end{verbatim}
\noindent

@@

This module defines any computation which outputs a Signal term. It comprises any sort of effect or filtering over a signal. It reproduces a real time processing by outputting processed frames as they are computed (non-deterministic call to a processing predicate) or stores the processed blocks to return and altered version of the input signal.

The module swiladspa offers communication with LADSPA which is the most generic and spread plugin API for general audio processing. The communication layer allows us to query the plugin description, control the set up of the plugin (control and audio ports) and retrieve the processed data. Through predefined routines we can offer a more handy access to the plugin or even declared predicates just defined by an input/output relationship.

@@

The WEKA API is interfaced through the package JPL available for SWI-Prolog defining a complex Java-Prolog interface to read/write ARFF files, load ARFF files, instantiate classifiers, execute classification and retrieve the classification scores of the dataset. This communication layer offers a really cumbersome and API-dependent module which requires semantics and interpretation rules to become into a useful logic-framework for audio data classification as described in the next section.

@@

So far we have presented how SWI-DSP provides a vocabulary of logic predicates to perform audio analysis computations, but this logic front-end is still inappropiate for such tasks. We have organized the modules in communication layers and task-oriented modules, but there is an important trade-off between understandability and specification. The communication layer even offering an application-independent access to the computation libraries still requires a high knowledge of such libraries besides a SWI-Prolog familiarization which is not expected in most part of the potential users. The task-oriented module, which is itself very layered, provides high-level predicates which are missing important options at setting-up.

In addition to this, we require a mechanism to update results and keep track of computation sequences, so that computations are not restricted to one SWI-Prolog session, allowing us to reuse and share them. This section tackles this problem by extending SWI-DSP as an expert system able to interpret data within defined ``DSP scenarios'' and by introducing Concurrent Transaction Logic on top of our extended predicate-logic language.

@@

'Classification'(Identifier, ListOfCategories, ListOfDataRecords, Classifier, ClassificationVector)

 \item Identifier gives a label for the classification set
 \item 'Category'(Label, Type, ValuesRange) specifies which type of data is going to be classified
 \item 'DataRecord'(RecordID, ListOfMatchingValues) is a data instance of the dataset 
 \item 'Classifier'(Name, ...)
 \item 'Score'(RecordID, ListOfValues) keeps track of each score (vector of scores depeding of the statistical distribution) for every record.
\end{itemize}

By defining these functors, the engine is able to interpret the data and the user has a clear semantic template to create the classification environment properly. We could even define some constraints and predefined functors instanes for a specific domain.

@@

We have presented a powerful reasoner for multimedia analysis implemented as a logic-based framework which allow us to convert low-level computations into shareable and distributed knowledge. The system design ''breaks apart`` traditional applications architecture to introduce logic and reasoning machinery on top of the computation engines providing a more intelligent, capable and understandable DSP reasoner. This reasoner can be extended and enriched by the open source APIs working on the foundations of the system and controlled by ''Prolog-speaking`` agents. SWI-DSP is fairly well positioned for the new web technologies and actually it finds in them the perfect semantic and storing support to overwhelm traditional and closed DSP systems.

Future work and development is oriented to:

\begin{itemize}
 \item Testing and comparison of performance with other rival systems
 \item Extending the SWI-DSP vocabulary and capabilities
\end{itemize}

SWI-DSP vocabulary and cabalities can be extended in several ways. First, by interfacing more open source APIs to cover the implementation of any DSP task (not necessarily restricted to audio domain). Secondly, by supporting symbolic data (MIDI) input which is a really important source of audio knowledge and providing analysis modules for such an input. Finally, other logic paradigms can contribute to enhance SWI-DSP (e.g. there are studies that use Inductive Logic Programming systems to extract harmonic rules or rhythm patterns given certain input descriptors [ref]). By providing semantics (OWL ontologies cleary point to this goal) to interconnect properly each logical module we can build a hierarchy of modules which help us to extract and share different levels of knowledge from digital signals. Such a system could be used either to extract low level cepstral coefficients from an input signal (mathematics level) or to obtain patterns of music (musicology level).

\begin{thebibliography}{4}

\bibitem{distributed} Raimond, Y., Sutton, C., Sandler, M.: A distributed data space for music-related information. In: Workshop on the Many Faces of Multimedia Semantic. ACM (Multimedia), Augsburg (2007).

\bibitem{ctl} Bonner, A.J., Kifer, M.: Concurrency and Communication in Transaction Logic. In: Joint International Conference and Symposium on Logic Programming, pp. 142--156. Bonn (1996)

\bibitem{tlprog} Bonner, A.J., Kifer, M.: Transaction Logic Programming. In: 10th International Conference on Logic Programming. Budapest (1993)

\bibitem{clam} CLAM, \url{http://www.clam.iua.upf.edu/}

\bibitem{marsyas} Music Analysis, Retrieval and Synthesis for Audio Signals. \url{http://marsyas.sness.net/}

\bibitem{jmir} McKay, C.: jMIR. \url{http://jmir.sourceforge.net/}

\bibitem{mo} Raimond, Y., Abdallah, S., Sandler, M., Frederick, G.: The Music Ontology. In: 8th International Conference of Music Information Retrieval. Vienna (2007)

\bibitem{kmrdf} Source code project of KM-RDF. \url{http://code.google.com/p/km-rdf/w/list}

\bibitem{ladspa} Linux Audio Developer Simple Plugin API. \url{http://www.ladspa.org/}

\bibitem{vamp} Cannam, C.: Vamp Plugins. \url{http://www.vamp-plugins.org/}

\bibitem{weka} WEKA home site \url{http://www.cs.waikato.ac.nz/ml/weka/}

\end{thebibliography}
